---
title: "Regression Models Course Project"
author: "Nadia Stavisky"
date: "1 September 2019"
output: pdf_document
---

```{r echo=FALSE}
#load libraries:
library(rmarkdown)
library(knitr)
library(ggplot2)        # plotting & data
library(GGally)
library(dplyr)          # data manipulation
library(tidyr)          # data re-shaping
library(plyr)
#install.packages("ggpmisc")
library(ggpmisc)
library(car)
library(magrittr)       # pipe operator
library(gridExtra)      # provides side-by-side plotting
```
```{r global.options, include = FALSE, echo=FALSE}
#set a golobal default options:
opts_chunk$set(
    echo = FALSE, # if FALSE knitr will not display code in the code chunk above it's results
    cache = FALSE, # if TRUE knitr will cache the results to reuse in future knits
    fig.width   = 5, # the width for plots created by code chunk
    fig.height  = 5, # the height for plots created by code chunk
    fig.align   = 'center', # how to align graphics in the final doc. 'left', 'right', 'center'
    fig.path    = 'figs/',  # file path to the directory where knitr shall store the graphics files
    #results     = 'asis',   # knitr will pass through results without reformatting them
    tidy = TRUE,
    tidy.opts=list(width.cutoff=60),
    strip.white = TRUE)     # if FALSE knitr will not remove white spaces at the beg or end of code chunk
```

##Introduction

The Motor Trend, a magazine about the automobile industry. Looking at a data set of a collection of cars, they are interested in exploring the relationship between a set of variables and miles per gallon (MPG) (outcome). They are particularly interested in the following two questions:

1. Is an automatic or manual transmission better for MPG?
2. Quantify the MPG difference between automatic and manual transmissions.


Motor Trend Car Road Tests - The data was extracted from the 1974 Motor Trend US magazine, and comprises fuel consumption and 10 aspects of automobile design and performance for 32 automobiles (1973-74 models).

#Source

Henderson and Velleman (1981), Building multiple regression models interactively. Biometrics, 37, 391-411.
A data frame with 32 observations on 11 (numeric) variables.

[, 1]	 mpg	 Miles/(US) gallon
[, 2]	 cyl	 Number of cylinders
[, 3]	 disp	 Displacement (cu.in.)
[, 4]	 hp	 Gross horsepower
[, 5]	 drat	 Rear axle ratio
[, 6]	 wt	 Weight (1000 lbs)
[, 7]	 qsec	 1/4 mile time
[, 8]	 vs	 Engine (0 = V-shaped, 1 = straight)
[, 9]	 am	 Transmission (0 = automatic, 1 = manual)
[,10]	 gear	 Number of forward gears
[,11]	 carb	 Number of carburetors

##Loading and preprocessing the data

```{r}
library(datasets)
data(mtcars)
?mtcars
```

#Exploratory data analysis

dataset structure:
```{r}
str(mtcars)
summary(mtcars)
```
There is no missing data in the data set.
Data type converted from numeric into factors for variables (cyl, vs, am, gear, carb).
Releveled variables:
-	 vs: Engine (0 = V-shaped, 1 = straight)
-	 am: Transmission (0 = automatic, 1 = manual)
```{r}
data <- mtcars %>% 
  mutate_at(vars(cyl, vs, am, gear, carb), ~ as.factor(.))
levels(data$am)
data$am <- revalue(data$am, c("0" = "automatic", "1" = "manual"))
data$vs <- revalue(data$vs, c("0" = "V-shaped", "1" = "straight"))
  head(data)
```
The variables included into dataset are:
```{r}
names(data)
```

The dataset contains `r dim(data)[1]` observation for `r dim(data)[2]` variables:
```{r}
dim(data)
```
We are interested in learning relationship beween mpg and other variables.
#Is an automatic or manual transmission better for MPG?

We will perform t-tests to compare two groups (car mpg performance for automatic and manual transmission type).
Note below the non-normality of the sample distribution which can be corrected with a log transformation.
MPG summary stats by transmission type (see Appendix 3):
- mpg summary for cars with automatic transmission:
```{r}
summary(data$mpg[which(data$am == "automatic")])
```
- mpg summary for cars with manual transmission:
```{r}
summary(data$mpg[which(data$am == "manual")])
```


Welch Two Sample t-test on if the automatic and manual averages differ, we are searching for any differences between the means rather than if one is specifically less than or greater than the other. 
```{r}
t.test(mpg ~ am, data = data)
```
The results of the t-test show p-value < 0.01 supporting the alternative hypothesis that "true difference in means is not equal to 0".

Alternatively, due to the non-normality concerns we can perform this test in two additional ways to ensure our results are not being biased due to assumption violations. We can perform the test with t.test and transform our data and we can also perform the nonparametric test with the wilcox.test function.
Welch Two Sample t-test with transformed data as log(mpg):
```{r}
t.test(log(mpg) ~ am, data = data)
```
Wilcoxon rank sum test with continuity correction:
```{r}
wilcox.test(mpg ~ am, data = data)
```
Both results support our initial conclusion that the Miles/(US) gallon of cars with automatic transmission is statistically different than the Miles/(US) gallon of cars with manual transmission.

#MPG difference between automatic and manual transmissions.

To quantify the MPG difference between automatic and manual transmissions let's explore the relationship between a set of variables and miles per gallon (MPG) (outcome):
- first we will regress Miles/(US) gallon on Transmission type (0 = automatic, 1 = manual) from our data set
```{r}
fitam <- lm(mpg ~ am, data = data)
summary(fitam)
```
Model summary that includs only transmission type gives us next performance mettrics:
1. Residual standard error: tells us that the actual mpg  deviate from the true regression line described in the model by approximately `r summary(fitam)$sigma`  Miles/gallon on average.
In the mtcars data set, the mean value of mpg is approximately `r  mean(data$mpg)` units, and so the percentage error is `r summary(fitam)$sigma/mean(data$mpg)*100`%
2. Multiple R-squared and Adjusted R-squared tells us proportion of variation in the mpg has been explained by this mode. 
3. F-statistic: 16.86 > 1, provides evidence against the null hypothesis H0 and suggests that at least one of the variables must be related to mpg.  
4. p-value: 0.000285  p-value associated with the F-statistic is much smaller then 0.05, we can infer that there is an association between the predictor and the response. We can reject the null hypothesis-that stands on there is no relationship to exist between X and Y. 

Model that includes only transmission type shows that on average mpg characteristic for cars with manual transmission higher on 7.2 Miles/(US) gallon without taking into account other characteristics.

Consider the correlation matrix for the quantitative predictor variables and response variable (Correlation matrix can be seen at Appendix1.):
```{r}
cor(mtcars)
```

From the correlation results we can see that the quantitative variables such as "hp" and "wt" are most correlated (negatively) with "mpg", also there is strong correlation between transmission type and some other variables such as cyl, disp, drat, wt, gear.

-Let's check affect of all other variables in the data set on mpg performance:
```{r}
fit1 <- lm(mpg ~., data = data)
summary(fit1)
```
Model summary that includs all variables gives us next performance mettrics:
1. Residual standard error: tells us that the actual mpg  deviate from the true regression line described in the model by approximately `r summary(fit1)$sigma`  Miles/gallon on average.
In the mtcars data set, the mean value of mpg is approximately `r  mean(data$mpg)` units, and so the percentage error is `r summary(fit1)$sigma/mean(data$mpg)*100`%
2. Multiple R-squared and Adjusted R-squared tells us proportion of variation in the mpg has been explained by this mode. 
3. F-statistic: 7.83 > 1, provides evidence against the null hypothesis H0 and suggests that at least one of the variables must be related to mpg.  
4. p-value: 0.000124  p-value associated with the F-statistic is much smaller then 0.05, we can infer that there is an association between the predictor and the response. We can reject the null hypothesis-that stands on there is no relationship to exist between X and Y. 

The coe???cient estimate for transmission type in the multiple regression model is much lomwer then in the first model, and the corresponding p-value is no longer signi???cant, with a value around 0.71. Does it mean that with concideration of other variables transmission type has no effect on mpg? To test if we can confirm that there is no relationship between transmission type and mpg, we will perform variable sellection procedure by choosing best model by AIC in a Stepwise Algorithm:
We will use [stepwise regression](https://en.wikipedia.org/wiki/Stepwise_regression) is a method of fitting regression models in which the choice of predictive variables is carried out by an automatic procedure.Backward elimination, which involves starting with all candidate variables, testing the deletion of each variable using a chosen model fit criterion, deleting the variable (if any) whose loss gives the most statistically insignificant deterioration of the model fit, and repeating this process until no further variables can be deleted without a statistically significant loss of fit.
Steps of the procedure for variable selection can be seen at Appendix2.
```{r}
best <- step(fit1, direction = "backward", trace = 0, steps = 1000)
summary(best)
```
Model summary that includs all variables gives us next performance mettrics:
1. Residual standard error: tells us that the actual mpg  deviate from the true regression line described in the model by approximately `r summary(best)$sigma`  Miles/gallon on average. the RSE decreased that proof the model improvement.
In the mtcars data set, the mean value of mpg is approximately `r  mean(data$mpg)` units, and so the percentage error is `r summary(best)$sigma/mean(data$mpg)*100`%
2. Multiple R-squared and Adjusted R-squared tells us proportion of variation in the mpg has been explained by this mode. The R-squared decreased due to the number of variables decresed. Adjusted R-squared increased the model with the largest adjusted R2 will have only correct variables and no noise variables.
3. F-statistic: 33.57 > 1, provides evidence against the null hypothesis H0 and suggests that at least one of the variables must be related to mpg.  
4. p-value: 1.506e-10  p-value associated with the F-statistic is much smaller then 0.05, we can infer that there is an association between the predictor and the response. We can reject the null hypothesis-that stands on there is no relationship to exist between X and Y. 

The final model include transmission type in the selected predictors, while it's p-value still >> 0.05.
Such effect can be explained with tendency (Notice the correlation between transmission type and other variables such as number of cylinders, weight, gear. Appendix 1.):
1. cars with automatic transmission more offen have higher number of cylinders then cars with manual transmission:
Distribution of cars by number of cylinders and transmission type (see Appendix 4):
```{r}
table(data$cyl,data$am)
```
2. cars with automatic transmission on average weight more then cars with manual transmission:
Weight summary by transmission type (see Appendix 5):
- weight summary for cars with automatic transmission:
```{r}
summary(data$wt[which(data$am == "automatic")])
```
- weight summary for cars with manual transmission:
```{r}
summary(data$wt[which(data$am == "manual")])
```
- Welch Two Sample t-test
```{r}
t.test(wt ~ am, data = data)
```
#Conclusion
On average, manual transmission is better than automatic transmission by 1.8 mpg. However, transmission type is not the only factor accounting for MPG, cylinders, horsepower, and weitght are the important factors in affecting the MPG.

#Appendix 1. Exploratory data analysis. Correlation between mtcars variables.
```{r fig1, fig.height = 35, fig.width = 35}
theme_set(theme_bw(base_size = 25))

ggpairs(data)
```
#Appendix 2. Stepwise variable selection.
```{r}
best <- step(fit1, direction = "backward", trace = 1, steps = 1000)
summary(best)
```

#Appendix 3. Exploratory data analysis. Correlation between mpg and am variables.
```{r fig2}
theme_set(theme_bw())
bp <- ggplot(data = data, aes(x=am, y=mpg)) +
  geom_boxplot(aes(fill = data$am), color = "grey30") +
  labs(title = "MPG by Transmission Type",x = element_blank(), fill ="Transmission Type") +
  theme(legend.position="bottom")
bp
```
```{r}
g <- ggplot(data, aes(mpg)) +
        geom_histogram(fill = "white", color = "grey30") +
        facet_wrap(~ am) +
  labs(title = "MPG histogram by Transmission Type",x = "MPG")
glog <- ggplot(data, aes(mpg)) +
        geom_histogram(fill = "white", color = "grey30") +
        facet_wrap(~ am) +
        scale_x_log10() +
  labs(title = "MPG histogram by Transmission Type",x = "Log(MPG)")
grid.arrange(g, glog, nrow = 2)
```
#Appendix 4.Exploratory data analysis. Distribution of cars by number of cylinders and transmission type.
```{r}
g <- ggplot(data, aes(x = data$cyl, fill = data$am)) + 
  geom_bar(stat="count",position="dodge", color = "grey30") +
  labs(title = "Count of cars by Number of cylinders by Transmission Type", x = element_blank(), fill ="Transmission Type") +
  theme(legend.position="bottom")
g
```
#Appendix 5. Exploratory data analysis. Weight summary by transmission type.
```{r}
g <- ggplot(data, aes(x = data$am, y = data$wt)) +
  geom_boxplot(aes(fill = data$am), color = "grey30")  +
  labs(title = "Weight by Transmission Type",x = element_blank(), fill ="Transmission Type") +
  theme(legend.position="bottom")
g

```
